---
title: "data.table"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data.table

> ###### resources:
- data.table [home page][dthp]
- data.table [cheat sheet][dtcc]

Data.table is enhanced data.frame.

From DataCamp blog: "Data.table allows you to reduce your programming time as well as your computing time considerably, and it is especially useful if you often find yourself working with large datasets.  For example, to read in a 20GB .csv file with 200 million rows and 16 columns, data.table only needs 8 minutes thanks to the fread() function.  This is instead of the hours it would take you with the read.csv() function. Once you understand its concepts and principles, the speed and simplicity of the package are astonishing!"

#### Creation of data.tables
- conversion from data.frame, matrix
- build form scratch in the same way as data.frame
- variables can be add using `:=` operator
- special characters, e.g. `.N` - length, `.` = `list`

**Example**:

```{r, include=TRUE}
dta = data.frame(DTM = seq(as.Date('1991-01-01'), as.Date('2000-12-31'), by = 'days'))
dta$X = rnorm(nrow(dta))
dta$GROUP = cut(dta$X, breaks = 4)
```        



```{r}
library(data.table)
dta = data.table(dta)
dta[, Y := rnorm(.N)]
```


#### General structure of data.tables calls:
`DT[i, j, by]` means: 

Take data.table `DT`, select rows using `i`, then calculate `j` possibly grouped by `by`. 

#### Selection

- the name of the data.table can be omitted, consider

```{r, eval = FALSE}
dta[DTM<as.Date('1992-01-01')]
dta[year(DTM) < 1993]
dta[X < 0, ]
dta[GROUP == GROUP[1]]
```

#### Column arithmetics

- the name of the data.table can be omitted, results either single vector, new variable inside data.table or new data.table

```{r, echo=TRUE, eval=FALSE}
# single vector
dta[, X * Y]
dta[, mean(X*Y)]

# new variable
dta[, Z := X * Y]
dta[, M := mean(X)]

# new data.table
dta[, .(X * Y, Z)]
dta[, .(XX = X * Y, Z)]

```

#### Calculation by groups

- groups are specified as the `by` argument

```{r, eval = FALSE}
dta[, mean(X), by = GROUP]
dta[, mean(X), by = month(DTM)]
```

- groups can be easilly combined

```{r, eval = FALSE}
dta[, mean(X), by = .(GROUP, month(DTM))]
```

#### Combination of data.tables

##### Concatenation

- by rows - `rbind` 
- by columns - `cbind` (in general not recommended - other database operations are prefered)

```{r}
a = data.table(x = 1:2, y = 1:2)
b = data.table(x = 3:4, y = c(3,1))
rbind(a, b)
```

##### merge

Two tables can be merged using common key columns (often date, station etc.). 

- general syntax:

```{r, eval=FALSE}
merge(x, y, by = NULL, by.x = NULL, by.y = NULL, all = FALSE, all.x = all, all.y = all)
```

**Examples**:

Key column same name, same coverage
```{r, eval = FALSE}
x = data.table(YEAR = 2005:2010, X = rnorm(6))
y = data.table(YEAR = 2005:2010, Y = rnorm(6))
merge(x, y, by = 'YEAR')
```

Key column different name, same coverage
```{r, eval = FALSE}
x = data.table(YEAR = 2005:2010, X = rnorm(6))
y = data.table(ROK = 2005:2010, Y = rnorm(6))
merge(x, y, by.x = 'YEAR', by.y = 'ROK')
```

Different coverage, outher join
```{r, eval = FALSE}
x = data.table(YEAR = 2000:2010, X = rnorm(11))
y = data.table(YEAR = 2005:2012, Y = rnorm(8))
merge(x, y, by = 'YEAR', all = TRUE)
```

Different coverage, inner join
```{r, eval = FALSE}
x = data.table(YEAR = 2000:2010, X = rnorm(11))
y = data.table(YEAR = 2005:2012, Y = rnorm(8))
merge(x, y, by = 'YEAR', all = FALSE)
```

Different coverage, left join
```{r, eval = FALSE}
x = data.table(YEAR = 2000:2010, X = rnorm(11))
y = data.table(YEAR = 2005:2012, Y = rnorm(8))
merge(x, y, by = 'YEAR', all.x = TRUE, all.y = FALSE)
```

Different coverage, right join
```{r, eval = FALSE}
x = data.table(YEAR = 2000:2010, X = rnorm(11))
y = data.table(YEAR = 2005:2012, Y = rnorm(8))
merge(x, y, by = 'YEAR', all.x = FALSE, all.y = TRUE)
```

##### Data.table fast join (data.table version >= v1.9.6)

- special data.table syntax:
```{r, eval=FALSE}
x = data.table(YEAR = 2000:2010, X = rnorm(11))
y = data.table(YEAR = 2005:2012, Y = rnorm(8))
x[y, on = 'YEAR', ]               # right join
x[y, on = 'YEAR', nomatch = 0]    # inner join
```



[dthp]: https://github.com/Rdatatable/data.table/wiki
[dtcc]: https://s3.amazonaws.com/assets.datacamp.com/img/blog/data+table+cheat+sheet.pdf                                                                                                                                                                                                                                                                                                                                                                                          