---
title: "R GIS"
output: 
  github_document: 
    toc: yes
---

<style type="text/css">
.table {

    width: 100%;

}
</style>

```{r, warning=FALSE, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '', prompt = TRUE)            
```


# Vector data

```{r, warning=FALSE}
library(sp) # provides classes
library(maptools) # provides i/o, coercion
```

### Classes

- `SpatialPoints`, `SpatialPointsDataFrame`
- `SpatialLines`, `SpatialLinesDataFrame`
- `SpatialPolygons`, `SpatialPolygonsDataFrame`
- `SpatialRing`, `SpatialRingDataFrame`
- `SpatialMultiPoints`, `SpatialMultiPointsDataFrame`
- `SpatialGrid`, `SpatialGridDataFrame`
- `SpatialPixels`, `SpatialPixelsDataFrame`

### Input/Output

```{r, eval = FALSE, tidy = TRUE}
# READING shapefiles
readShapePoints(fn)
readShapeLines(fn)
readShapePoly(fn)
readShapeSpatial(fn)

# WRITTING shapefiles
writePointsShape(x, fn)
writeLinesShape(x, fn)
writePolygonsShape(x, fn)
writeSpatialShape(x, fn)

# more general - incl. geodatabases etc.
library(rgdal)
ogrDrivers() # list supported drivers
ogrListLayers(dsn) # list layers from Data Source Name (dsn)
ogrInfo(dsn, layer) # layer info
ogrFIDs() # list field IDs (FIDs)
OGRSpatialRef() # projection (proj4string)
readOGR(dsn, layer) # reading (reads also projection)
writeOGR(dsn, layer) # writting (writes also projection)
```

### Data and metadata access

*Creating toy data - `SpatialPointsDataFrame`*

```{r, include = TRUE, echo = -1}
set.seed(13131)
d = data.frame(ID = sample(LETTERS, 10), x = rnorm(10), y = rnorm(10), ELE = 300 + rnorm(10, 20, 6)^2)
row.names(d) = d$ID
p = SpatialPointsDataFrame(d[, 2:3], d)
```

###### Number of objects
```{r}
length(p)
```

###### Getting names of variables from attribute table (for `Spatial*DataFrame` only)
```{r}
names(p)
```

###### Accessing attribute table
```{r}
p@data
```

###### Accessing individual variables
```{r, eval = FALSE}
# by name
p$ID 
p@data$ID 
p@data[, 'ID']

# by position
p@data[, 3]
```


###### Selecting individual objects
```{r, eval = FALSE}
p[3:4, ] # by position
p['B', ] # by name
p[p$ELE < 500, ] # by condition
```

### Spatial analysis

```{r message=FALSE, warning=FALSE}
library(rgeos) # provides many functions
```

command  | meaning
------------------------------ | ----------------------------------------------------------------------------
`gBuffer(spgeom, byid, width)` | creates buffer of specified width, possibly separately for each geometry (when `by = TRUE`)
`gCentroid` | get centroid
`gContains`, `gCovers` | geometry relationships
`gDelaunayTriangulation` | triangulation
`gDifference` | geometry difference
`gDistance` | distance in map units
`gIntersection` | find intersection of two spgeoms
`gIntersects(spgeom1, spgeom2)` | does spgeom1 intersect spegeom2 
`gLength` | geometry length
`gSimplify` | simplify geometry
`gUnion`, `gUnionCascaded` | union geometries
`gWithinDistance` | returns `TRUE` if the distance of two geometries is less then specified distance

# Raster data

```{r, message=FALSE}
library(raster)
```

### Classes

class  | 
--------- | ---------------------------------------------------------------------
`raster` | raster layer
`brick`  | in-memory multi/layer raster object
`stack`  | collection of raster layers with the same spatial extent and resolution


### Input/output
```{r, eval = FALSE}
# reading rasters
raster(file)
brick(list(file1, file2, ...))
stack(list(file1, file2, ...))

# writting rasters
writeRaster(r, 'file')
```

### Manipulation

###### Working with layers
```{r, eval = FALSE}
# create brick from 2 layers
b = brick(layer1, layer2)
# select first layer
b[[1]]
```

###### Getting and setting values

```{r, eval = FALSE}
getValues(x) # extract values
values(x) # the same

setValues(x, values) # set values
values(x) = values # the same
```

###### Raster calculations
```{r, eval = FALSE}
x + y # sum the values of raster x and raster y
x / y # etc. ...
mean(x) # mean of x across layers

calc(x, fun) # apply fun to values of raster x
cellStats(x, fun) # compute statistics for the cells of each layer of a Raster* object
```

###### Modification of raster objects
```{r, eval = FALSE}
resample(x, y) # transfer values from raster x to raster y (with same projection but different resolution)
projectRaster(x, y) # project raster x to projection and resolution of raster y
aggregate(x, fact) # aggregate values to lower resolution
disaggregate(x, fact) # create raster with higher resolution

crop(x, s) # crop raster x by spatial object s
mask(x, s) # mask values of raster x by spatial object s
extract(x, s, fun) # extract values from raster r according to spatial object s, optionally using aggregate function fun
```

# Coordinate reference systems

- specified as "proj4string"

###### Set or get projection
```{r, eval = FALSE}
# getting projection
projection(x)
crs(temp)
x@crs # for raster class
x@proj4string # for sp* classes

# setting proj4string (this is different from projecting raster)
projection(x) = crsString
crs(x) = crsString
```

###### Transformations
```{r, eval = FALSE}
# vectors
spTransform(x, CRS) # transform x (with specified CRS) to new CRS

# raster
projectRaster(r, crs = CRS) # project raster r to new CRS
```

###### Specifying projections
```{r}
# WGS
CRS('+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs')
# krovak S-JTSK
CRS('+proj=krovak +lat_0=49.5 +lon_0=24.83333333333333 +alpha=0 +k=0.9999 +x_0=0 +y_0=0 +ellps=bessel +units=m +no_defs')
# specification by epsg
CRS('+init=epsg:4326') # WGS
CRS('+init=epsg:2065') # krovak
```

***